Still Missing a Cool Title

### 2006-05-31 talk/dls2006/paper.tex
Move the footnote numbers after the "," or "." or ":".

The front-end part of the translation process analyses the input
RPython program in two phases, as follows\footnote{Note that the two
phases are intermingled in time, because type inference proceeds from
an entry point function and follows all calls, and thus only gradually
discovers (the reachable parts of) the input program.}:

@@

  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
             then a stack of transformations
  ]
\end{verbatim}
%
\begin{enumerate}
\item We take as input RPython functions\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This allows us to use unrestricted
       python for meta-programming purposes at load time, in a
       seemingly staged programming approach, in which the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.}, and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

@@

For example, \texttt{ll\textunderscore{}append()} in figure \ref{fig_llappend}
is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.}, which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g.\ Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}.

@@

In conjunction with the Flow Object Space, the bytecode interpreter of
PyPy thus performs abstract interpretation of Python bytecodes from
the application\footnote{Note that this process uses the
\textit{unmodified} bytecode interpreter.  This means that it is
independent of most language details.  Changes in syntax or in
bytecode format or opcode semantics only need to be implemented once,
in the bytecode interpreter.  In effect, the Flow Object Space enables
an interpreter for \textit{any} language to work as a front-end for
the rest of the tool-chain.}.  In this case, the bytecodes in
question come from the RPython application that we would like to
translate.

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later.  Moreover, we rely on this feature to perform
compile-time computations, particularly in the generic system code
helpers, which ask for and compute with the concrete types of the
variables they receive.}

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment:
%
\begin{verbatim}
    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)
\end{verbatim}
%
Although the type of \texttt{x} may be some parent class of
\texttt{MyClass}, it can be deduced to be of the more precise type
$Inst(MyClass)$ within the positive branch of the \texttt{if}.
\footnote{Remember that our graphs are in SSI form, which means that the
\texttt{x} inside each basic block is a different Variable with a
possibly different type as annotation.  Given that our type inference
is not flow-sensitive, SSI gives an advantage over SSA here.}

>>>>

The front-end part of the translation process analyses the input
RPython program in two phases, as follows:\footnote{Note that the two
phases are intermingled in time, because type inference proceeds from
an entry point function and follows all calls, and thus only gradually
discovers (the reachable parts of) the input program.}

@@

  [figure 0: flow graph and annotator, e.g. part of doc/image/translation.*
             then a stack of transformations
  ]
\end{verbatim}
%
\begin{enumerate}
\item We take as input RPython functions,\footnote{The input to our
       translation chain are indeed loaded runtime function objects,
       not source code or ASTs.  This allows us to use unrestricted
       python for meta-programming purposes at load time, in a
       seemingly staged programming approach, in which the whole of
       the source program -- as Python program -- produces the RPython
       program input to the tool-chain as the object graph loaded in
       memory.  This includes both the relevant functions and prebuilt
       data.} and convert them to control flow graphs -- a structure
       amenable to analysis.  These flow graphs contain polymorphic
       operations only: in Python, almost all operations are
       dynamically overloaded by type.

@@

For example, \texttt{ll\textunderscore{}append()} in figure \ref{fig_llappend}
is a Python function
that manipulates objects that behave like C structures and arrays.
This function is inserted by the LLTyper, as a helper to implement the
\texttt{list.append()} calls found in its RPython-level input graphs.
By going through the front-end reconfigured to use C-level types, the
above function becomes a graph with such C-level types,\footnote{The
low-level type system specifies that the function should be
specialized by the C-level type of its input arguments, so it actually
turns into one graph per list type -- list of integers, list of
pointers, etc.  This behavior gives the programmer a feeling
comparable to C++ templates, without the declarations.} which is then
indistinguishable from the other graphs of the forest produced by the
LLTyper.

@@

Note that such system code cannot typically be expressed as normal
RPython functions, because it corresponds to primitive operations at
that level.  As an aside, let us remark that the number of primitive
operations at RPython level is, comparatively speaking, quite large:
all list and dictionary operations, instance and class attribute
accesses, many string processing methods, a good subset of all Python
built-in functions...  Compared to other approaches [e.g.\ Squeak], we
do not try to minimize the number of primitives -- at least not at the
source level.  It is fine to have many primitives at any high enough
level, because they can all be implemented at the next lower level in
a way that makes sense to that level.  The key reason why this is not
burdensome is that the lower level implementations are also written in
Python -- with the only difference that they use (and have to be
typeable in) the lower-level type system.\footnote{This is not strictly
true: the type systems are even allowed to co-exist in the same
function.  The operations involving higher-level type systems are
turned into lower-level operations by the previous transformations in
the chain, which leave the already-low-level operations untouched.}

@@

In conjunction with the Flow Object Space, the bytecode interpreter of
PyPy thus performs abstract interpretation of Python bytecodes from
the application.\footnote{Note that this process uses the
\textit{unmodified} bytecode interpreter.  This means that it is
independent of most language details.  Changes in syntax or in
bytecode format or opcode semantics only need to be implemented once,
in the bytecode interpreter.  In effect, the Flow Object Space enables
an interpreter for \textit{any} language to work as a front-end for
the rest of the tool-chain.}  In this case, the bytecodes in
question come from the RPython application that we would like to
translate.

The Flow Object Space records all operations that the bytecode
interpreter "would like" to do between the placeholder objects.  It
records them into basic block objects that will eventually be part of
the control flow graph of the whole function.  The recorded operations
take Variables and Constants as argument, and produce new Variables as
results.  The Constants serve two purposes: they are a way to
introduce constant values into the flow graphs -- these values may be
arbitrarily complex objects, not just primitives -- and they allow
basic constant propagation.\footnote{This is useful at this level for
some constructs of the bytecode interpreter, which can temporarily
wrap internal values and push them onto the regular value stack among
the other application-level objects.  We need to be able to unwrap
them again later.  Moreover, we rely on this feature to perform
compile-time computations, particularly in the generic system code
helpers, which ask for and compute with the concrete types of the
variables they receive.}

@@

Of course, this would be pointless if the annotation did not give
precise enough information for our needs.  We must describe a detail of
the abstract interpretation engine that is critical for precision: the
propagation of conditional types.  Consider the following source code
fragment:
%
\begin{verbatim}
    if isinstance(x, MyClass):
        f(x)
    else:
        g(x)
\end{verbatim}
%
Although the type of \texttt{x} may be some parent class of
\texttt{MyClass}, it can be deduced to be of the more precise type
$Inst(MyClass)$ within the positive branch of the
\texttt{if}.\footnote{Remember that our graphs are in SSI form,
which means that the
\texttt{x} inside each basic block is a different Variable with a
possibly different type as annotation.  Given that our type inference
is not flow-sensitive, SSI gives an advantage over SSA here.}
